The weight of the mountains: Constraints on tectonic stress, friction, and fluid pressure in the 2008 Wenchuan earthquake from estimates of topographic loading

### 2014-09-17 revisions/rebuttal.md

## R1a
> The complete stress tensor is divided into contributions from topography,
> tectonic, and lithostatic stress (eq 9). The tectonic stress is then assumed
> to increase linearly with depth "so that the entire crust is near the critical
> failure envelope" (line 247), referencing Towend and Zoback 2000. This is
> essentially just a statement of Byerlee's law, in which brittle failure occurs
> in the upper crust at a stress level approximately equal to the lithostatic
> pressure applied to the rock scaled by ~0.6 (this study's parameter T'?).

This is related to, but not really a restatement of, Byerlee's law. This is
much more a statement of the Coulomb failure envelope. Byerlee's law is related
to the failure of intact rock, which is often quoted (as in the review, and
in our section 6.3) as 0.6. This coefficient is related to the coefficient of
internal friction of intact rock, which is not our $T'$. In our study, we are
primarily concerned with the failure of a specific, pre-existing fault, whose
strength (coefficient of static friction) is unknown but likely less than
unfractured rock. We scale tectonic stresses linearly with depth, and each
component of tectonic stress ($T_{xx}$, $T_{xy}$, and $T_{yy}$, or $T_{max}$,
$T_{min}$, and azimuth) increases with depth at a different rate.

$T'$ is simply $T / \rho g z$ (line 275 of original manuscript).

## R1b
> However, this means that the formulation expressed in equation 9 is counting
> the lithostatic contribution of stress twice.

We have not double-counted lithostatic stress, we have simply been unclear in
our description of the stress state. Lithostatic stress is an isotropic tensor,
so (as described) it is only the diagonals of the tensor. Tectonic stress is
only horizontal, $T'$ is scaled by $\rho g z$ (not by the tensor $L$, as we had
previously, and erroneously stated). So $T$ and $L$ do not occupy the all same
components of the stress tensor.

We amend the submitted description:

> We consider the complete stress tensor, $S$, at a point in the crust to
> be
> 
> \begin{equation}
> S = M + T + L \; ,
> \end{equation}
> 
> where $M$ is described above, $T$ is the tectonic stress tensor, and $L$
> is the lithostatic stress tensor. $L$ is isotropic, with diagonal
> components equal to $\rho g z$. We assume that $T$ is laterally
> homogeneous and only has horizontal stress components (i.e.,
> $T_{xz} = T_{yz} = T_{zz} = 0$, with $T_{xx}$, $T_{yy}$ and $T_{xy}$
> non-zero). We further assume that $T$ increases linearly with depth so
> that the entire upper crust is near the critical failure envelope
> [e.g., *Townend and Zoback, 2000*], and thus we parameterize the components of $T$
> as scalars multiplied by lithostatic pressure, denoted as $T^\prime$

> We consider the complete stress tensor, $S$, at a point in the crust to
> be

> \begin{equation}
> S = M + T + L \; ,
> \end{equation}

> where $M$ is described above, $T$ is the tectonic stress tensor, and $L$
> is the lithostatic stress tensor. $L$ is isotropic, with diagonal
> components equal to $\rho g z$. We assume that $T$ is laterally
> homogeneous and only has horizontal stress components (i.e.,
> $T_{xz} = T_{yz} = T_{zz} = 0$, with $T_{xx}$, $T_{yy}$ and $T_{xy}$
> non-zero). The expanded stress tensor is:

> \begin{equation}
> S = \begin{bmatrix}
>   M_{xx} + T_{xx} + L_{xx} & M_{xy} + T_{xy} &  M_{xz} \\
> 	M_{xy} + T_{xy} &  M_{yy} + T_{yy} + L_{yy} & M_{yz} \\
> 	M_{xz}     &  M_{yz}  &  M_{zz} + L_{zz}
> 	\end{bmatrix} \; .
> \label{eqn:stress_tensor}
> \end{equation}

> We further assume that $T$ increases linearly with depth so
> that the entire upper crust is near the critical failure envelope
> [e.g., *Townend and Zoback, 2000*], and thus we parameterize the components 
> of $T$ as scalars multiplied by $\rho g z$, denoted as $T^\prime$. Therefore, 
> if $T'_{xx} = 0.1$, at some point just below 1 km, $L=27$ MPa, so
> $S_{xx} = M_{xx}$ + 27 MPa + 2.7 MPa.

which hopefully clarifies this.

## R1c
> The Towend paper derives its results from deep drilling studies, so it
> would be more appropriate to assume that the total stress S increases 
> linearly with depth in this manner.

We agree that $S$ increasing linearly with depth would be more analogous to
*Townend and Zoback 2000*, but our model does not allow for this, because we
would have to vary $T$ spatially to exactly counterbalance variations in $M$
that deviate from linearity. Instead, we find it much simpler to approximate
$T$ as increasing linearly with depth, because then we can quantify it with
three numbers ($T'_{xx}$, $T'_{xy}$, and $T'_{yy}$, or equivalently $T'_{max}$,
$T'_{min}$ and the azimuth of $T'_{max}$). Additionally, because $L$ increases
rapidly with depth, it overwhelms the largest topographic stresses on the
fault (~80 MPa) at about 3 km, so for most of the crust, $S$ does
increase essentially linearly with depth.

@@

We agree that $S$ increasing linearly with depth would be more analogous to
*Townend and Zoback 2000*, but our model does not allow for this, because we
would have to vary $T$ spatially to exactly counterbalance variations in $M$
that deviate from linearity. Instead, we find it much simpler to approximate
$T$ as increasing linearly with depth, because then we can quantify it with
three numbers ($T'_{xx}$, $T'_{xy}$, and $T'_{yy}$, or equivalently $T'_{max}$,
$T'_{min}$ and the azimuth of $T'_{max}$). Additionally, because $L$ increases
rapidly with depth, it overwhelms the largest topographic stresses on the
fault (~80 MPa) at about 3 km, so for most of the crust, $S$ does
increase essentially linearly with depth.

## R2
> The authors make the "Wallace-Bott" assumption (line 277) that fault slip
> direction aligns with maximum horizontal principal stress direction, and the
> subsequent analysis of tectonic stress is made with the criteria that
> acceptable solutions minimize the difference between these two (equation 11
> and lines 300-301). However, several of the results described in discussion
> section 6.2 interpret the alignment of the modeled tectonic stress
> orientation and coseismic slip direction as a meaningful result: e.g., line
> 401, line 407, line 417. For example, that the shear stress has non-trivial
> contributions of both strike and dip-slip was required since the stress must
> align with the oblique-slip earthquake.

In these comments, the reviewer basically asserts that we set up our model to
replicate observations and then claim that is significant that our model's
predictions match the observations, and then gives several example lines (401,
407, 417). Though this is not the interpretation of the paragraph that we had
intended, it is apparent how this interpretation could be made. Nonetheless,
the specific lines seem to be off the mark in this:

Line 401 is basically a quick description of the results (E-W maximum principal
tectonic stress produces oblique shear on the fault), and is not a claim about
the significance of the results.

Line 407 is a comparison of our estimated maximum principal stress and stress
measurements from the WFSD drill hole at that location; the drill hole stress
measurements were not used in our inversion, and therefore this is an
independent confirmation of our results, not a direct consequence of the
inversion.

@@

to

> This stress configuration is in close agreement with pre-earthquake stress
> measurements near the rupture zone...

## R3
> Line 188: In the Liu and Zoback formulation, tectonic stress is an integrated
> part of the horizontal loading functions (equations 5 and 6). Is it 
> appropriate to neglect tectonic stress in these equations but then add it 
> back in linearly during the analysis of section 4? I expect this has some
> ramifications.

This was something we puzzled over for a while. Ultimately, we chose our
approach because it is less than obvious how to incorporate a depth-dependent
tectonic stress into the calculations. The horizontal loading functions specify
horizontal surface loads that are then propagated into the subsurface using
Cerruti's solutions [Appendix A2]. The Liu and Zoback formulations of the
loading functions require a fixed tectonic stress, or the tectonic stress at
zero depth. Because we assume that tectonic stress increases linearly from zero
with depth, the tectonic stress at the surface is zero.

An additional complication is that the calculation of topographic stresses in
the region (not their interpolation onto the fault) takes about 9 hours, and
so devising some iterative scheme where we could calculate topographic
stresses, then find tectonic stress and then add that back into the topographic
stress calculations and repeat until some criteria is reached, is really out
of the question with the Bayesian inversion scheme we've chosen (there are
about 18,000 models in the joint posterior).

@@

## R4
> Line 193 assumes a "Poisson halfspace", presumably meaning an elastic
> halfspace made up of a Poisson solid with Poisson ratio of 0.25. While this
> is an oft-made assumption, in this situation it has more impact. Figure 2
> nicely illustrates how the topography loading on the fault planes can change
> sign depending on how effective the vertical load is at generating horizontal
> stresses, which is precisely what the Poisson ratio governs. It would be
> helpful, therefore, to know how robust these results are to choice of
> Poisson ratio.

We had chosen our the value of 0.25 for the Poisson ratio not only because of
the calculational convenience (the shear modulus and Lame's parameters drop out
of the equations) but because this is the most likely value for the rather
felsic Tibetan crust based on seismic studies.  But in light of these comments,
we have recalculated the fault stresses using a Poisson ratio of 0.28, as a
test. We have found that the stresses are different by 1 or 2 MPa, which is
a very small difference relative to the overall magnitudes. We have changed the
relevant sentence in first paragraph of Section 2.2 from:

> We assume a Poisson halfspace

> We assume a Poisson ratio of 0.25, following receiver function studies
suggesting values of ca. 0.24-0.26 throughout central and western China [*Chen
et al., 2010*] (we have tested a Poisson ratio of 0.28, which is on the higher
end of values for intermediate rock compositions [*Zandt and Ammons, 1995*],
and found the results to vary by a few percent).

@@

> We assume a Poisson ratio of 0.25, following receiver function studies
suggesting values of ca. 0.24-0.26 throughout central and western China [*Chen
et al., 2010*] (we have tested a Poisson ratio of 0.28, which is on the higher
end of values for intermediate rock compositions [*Zandt and Ammons, 1995*],
and found the results to vary by a few percent).

## R5
> Line 292: do both the tectonic stresses have to be positive? Vertical 
> principal stress could still be the least compressive if Tmin were negative 
> and Tmax were positive, right? This would increase the available permitted 
> parameter space.

In general, neither of the tectonic stresses have to be positive. However,
vertical stress cannot be least compressive if $T_{min}$ were negative
(tensile), unless topographic stress produced a strong horizontal compression
in the direction of $T_{min}$ and less strong vertical compression. Our results
show that the topographic stresses are mostly stronger (more compressive) in
the vertical direction than in the horizontal directions.

Additionally, from a tectonic standpoint, there is no real reason to consider
N-S tectonic tension (the direction of $T_{min}$ in our results) in the Sichuan Basin;
this is a region surrounded by thrust and strike-slip faults of various
orientations (e.g. Figure 1). If there was tectonic tension in the region, then
the patterns of deformation would be different in the interior of the Sichuan
Basin away from topography; currently faulting there is thrusting, which
indicates a vertical $\sigma^3$. There is no evidence of N-S extension anywhere
in the Himalayan-Tibetan orogen except for in the Miocene in the Himalaya,
which is synchronous with parallel thrusting (the famed Himalayan extrusion).

The purpose of Bayesian priors is to keep likely values in as much as to keep
unlikely values out. Because we adopt a relatively simple scheme of uniform
probabilities over some interval, we can't add unlikely values at the same
rate as more likely values, without either strongly diluting the number of
acceptable models in the posterior (and therefore needing to run many more
models, and we are already at the limit of what can be done with 128 GB RAM),
and/or introducing a lot of unlikely models into the posterior, depending on
how sensitive the results are to the priors.

@@

## R6
> The emphasis on the Bayesian estimation providing a doubly bounded constraint 
> on tectonic stress (379) or on the on the ability of the study to identify 
> more than a minimum bound of the tectonic stress (line 373) is overstated. 
> The fundamental gist of this study is that topography provides a resistive
> force that must be overcome by tectonic stress (lines 140-142). This provides
> a physical minimum bound for tectonic stress, but there is no information
> considered that physically provides any upper bound. Unsurprisingly,
> therefore, the results in figure 6 show a pdf with a lower bound only, as
> noted on line 413. This is not a failure or triumph of the chosen statistical
> scheme, it's a physical limitation of the available data.

While the results are certainly more sensitive to the low end of allowable
$T'_{max}$ values, they are somewhat sensitive to the high end. This
sensitivity varies with slip model, as is apparent in Figure 6. Figure 8,
the joint results, show that $T'_{max}$ decreases past about 1.0, and the
likelihood of $T'_{max}=2.5$ is about 60% of the likelihood of $T'_{max}=0.8$.
Additionally, this is assuming that $T'_{max}$ is *a priori* equal at 0.8 and
2.5 (which is our simple model); personally, I find it rather unlikely that
tectonic stresses would be so high, although we chose the model prior to be
a bit different than our personal prior beliefs.

@@

The lines 373 and 379, chosen by the reviewer, are in a discussion of
methodological differences between this study and another that only attempts to
constrain the minimum tectonic stress in a similar environment; the differences
here are in approach more than results.

## R7
> I'm concerned that the Bayesian PDF estimation as used for T'min, e.g. in 
> figure 6, results in a biased estimate. Because there's a requirement that 
> Tmin < Tmax, half of parameter space is off limits a priori (lines 293-296).
> It therefore seems unsurprising that the T'min PDF is heavily weighted toward
> zero. Is there a way to re-scale the problem so that the parameters being
> estimated are independent?

The requirement that $T_{min} < T_{max}$ is not simply an attribute of our
priors or Bayesian estimation scheme, but basic math and logic. Minimum is
defined to be less than maximum, so the half of the parameter space where this
would occur is off limits by definition.

Constructing suitable priors, though, was something that we spent some time on.
The issue with picking stresses independently is that they're not independent,
because $T_{min} < T_{max}$, regardless of the value of $T_{max}$, so there is
an inherent conditional probability. So they can't be sampled indendently, and
there are two approaches that can be taken. One of them is to take a random
(uniform) Nx2 array and sort each row, so that one column is $T_{max}$ and the
other is $T_{min}$. This approach yields a uniform joint prior probability
density over the available half of the $T_{min}$ x $T_{max}$ parameter space,
but yields marginal priors for $T_{max}$ that are strongly biased towards high
values, and priors for $T_{min}$ that are biased towards the low end.

The other is to do what we did, and take a random (uniform) Nx1 array for
T$_{max}$, and then take $T_{min}$ as some random value from [0,1) of
$T_{max}$. This gives a uniform prior PDF for $T_{max}$, a similar prior for
$T_{min}$ as the first option, and a higher joint probability density towards
the low $T_{max}$ end of the $T_{max}$ vs $T_{min}$ space.

So, although the joint probability density is higher towards the low $T_{max}$
end (visible in Figures 6 and 8a) in our approach, the marginal priors for
$T_{max}$ would be much more strongly biased in the first approach, but
$T_{min}$ is less affected by the sampling strategy. Overall, we feel that our
approach is less biased, because it results in a uniform $T_{max}$, which is
the more important value.

@@

On further reflection, it would be possible to pick a uniform $T_{min}$ and
then build $T_{max}$ around that, but then $T_{max}$ would be non-uniform. We
can think of no particular reason to favor this approach, since researchers are
generally more interested in estimating maximum principal tectonic stress than
minimum.

However, it is clear in Figure 8b, where $T'_{min}$ is scaled to $T'_{max}$,
that the *ratio* of $T_{min}$ to $T_{max}$ is constrained by the inversion;
the most likely value for $T_{min} \approx 0.3 T_{max}$, regardless of the
value for $T_{max}$. This point was mentioned in passing in section 5.1 (line
330) but could be discussed in more detail, particularly with reference to 
Figure 8.1. Therefore, we add to Section 5.2:

>It is also apparent in Figure 8b that regardless of the magnitude of
$T^\_{\mathrm{max}$, $p_{J}(T^\prime_{\mathrm{min}} | D)$ is
consistently 0--0.6 $p_{J}(T^\prime_{\mathrm{max}} | D)$, with a mode
of about 0.3; this indicates that the relative magnitude of the
tectonic stresses has a substantial influence on the rake of the maximum
shear stress resolved on the fault.

@@

## R8
> The caption for Figure 4 is truncated on p 43, and is not present anywhere
> else in the text. The figure itself is not self-explanatory, and the details
> aren't sufficiently explained in the results section 3. This seems to be an
> unfortunate LaTEX mishap, but since this model is the primary finding of this
> study and the basis for the subsequent analysis, the reader unfortunately
> cannot just "read past" the omitted details. Some of the obvious gaps are
> noted below.

This is indeed true, and it is unfortunate. However, the unclipped figure
caption was available to the reviewer in the 'article file' available to
him/her (not the merged pdf) on the review website. More fortunately, the
general reader will be reading a professionally typeset version of the paper,
not a submitted LaTeX draft.

@@

## R9
> It is not entirely clear which faults are shown in 4a, and not at all clear
> what faults or regions of space are represented in 4b-c (by the way, a b and
> c are not labeled either). What depths are represented in these models? The
> visible portion of the caption notes "same lateral extent, but different
> perspectives", but it is not clear what this means. And why is the maximum
> slip magnitude in "A" 7m, but the maximum component slips in "B" and "C" are
> 14m and 10m?

The caption states that

The magnitude of slip in the models varies by a factor of about 2. The rakes
and the spatial distribution of slip vary much less.

@@

## R10
> Also, as section 3 represents the principal findings of this study, the 
> reader would benefit from some additional discussion here of the relevant
> features of the models. For example, the text on lines 237-239 and lines
> 244-247 seem to represent and important result worthy of more attention and
> discussion.

We have added a paragraph and a half describing the characteristics of the slip
models used, including the similarities and differences of their geometry and
slip distributions, to Section 2.3:

>We use six models,
those of *Shen, et al.* [*2009*], *Feng, et al.* [*2010*], *Zhang, et al.*
[*2011*], *Fielding, et al.,* [*2013*] and two from *Qi, et al.* [*2011*] (these two models share a
fault geometry but use different smoothing functions for the slip
distribution). These vary somewhat in the fault geometry and specifics of the
slip distribution, but nonetheless are of high resolution, are properly
georeferenced. By using a suite of models in our calculations instead of a
single model, we evaluate results that are persistent in most or all of the
models to be more robust, and other results can be more confidently linked to
specifics of the model geometry or slip distribution.

### 2014-09-17 revisions/styron_hetland_wenchuan_topo_revised.tex
revisions before sending to eric

\begin{article} \section{Introduction}\label{introduction}

\begin{quote} ``And He has set up on the earth mountains standing firm, lest it
    should shake with you'' -- \emph{The Holy Qur'an}, 15:16 \end{quote}

Stress is of fundamental importance to many processes in the earth. Both the
isotropic and deviatioric components of stress exert control on the deformation
state of the earth at any point in the brittle regime. However, unlike other
fundamental quantities such as temperature, stress is typically difficult to
measure \emph{in situ}, without drilling-based techniques. Therefore, stress is
often treated in a semi-quantitative manner, with an emphasis on directions and
relative magnitudes of the principal stresses, either locally or regionally
\citep[e.g.,][]{angelier1994}. These estimates of stress are commonly derived
from strain, for example from studies of earthquake focal mechanisms
\citep[e.g.,][]{michael1987} or of fault slip data \citep[e.g.,][]{reches1987,
medinaluna2013}.

In areas of substantial relief, high terrain and steep slopes generate large
stresses in the crust beneath and adjacent to the high topography
\citep{jeffreys1924, coblentz1996}. Because of the irregularity of topography
in mountainous regions, the stresses produced by topography are also
heterogeneous, and may play a prominent role in local or regional deformation,
particularly if the region is tectonically active.  For example, shear and
normal stresses on a fault from topographic loading may push a particular fault
closer to or (as suggested by the Qur'an) farther from failure, or reorient the
net shear stress direction on a fault. These effects may affect the
localization or deformational style in a region. Furthermore, heterogeneous
topographic stresses on a particular fault may affect the way earthquake
ruptures propagate across the fault plane. Despite this, the degree to which
topographic stresses affect faulting has received little direct study.

@@

Aspects of topographic stresses and their relevance to tectonics have been
studied for some time. \citet{jeffreys1924} noted that the presence of high
mountains is evidence that the Earth's crust can support significant
heterogeneous differential stresses over long time periods.
\citet{dalmayrac1981} and \citet{molnar1988} discussed how extensional
deformation in the high parts of orogens that is temporally coincident with
contractional deformation at the low-elevation margins of the orogen may be
explained by a spatially invariant, depth-integrated horizontal compressive
stress with spatially varying vertical stresses caused by topography.
\citet{richardson1994} exploited this relationship in the central Andes to
estimate horizontal tectonic stresses through finite element modeling, and more
recently \citet{copley2009} performed similar work in Albania.
\citet{bollinger2004} showed how increased normal stress on the Main Himalayan
thrust due to loading of the high Himalayan massifs locally suppressed
microseismicity and increased fault locking. Meade and Conrad \citet{meade2008}
demonstrated that the increased weight of the uplifting Andes influenced the
Nazca-South America convergence rate.

@@

Topography was taken from the CGIAR-CSI v.4 release
\citep{jarvis2008srtm} of the Shuttle Radar Topographic Mission
\citep{farr2007srtm} Digital Elevation Model (DEM) at 1 km nominal
resolution. The DEM was projected from native WGS84 geographic
coordinates to UTM zone 48N, decreasing the nominal horizontal
resolution to 851 m. We assume a Poisson halfspace, and Green's
functions for the Boussinesq and Cerruti point-source solutions were
calculated at regular points in large 2-D grid at each depth
considered, with the point-source centered in the grid (see Table
\ref{table:convo_params} for model parameters). A mask was applied to
each of the discretized Green's functions such that values outside a 
radius (i.e. the `corners' of the array) were set to zero, yielding a
circular array. The size of the
grid was chosen to be quite large to incorporate potential contributions
from the elevated topography throughout eastern Tibet. So that the
Green's functions and the topography were discretized on the same size
grid, we pad the Green's function array with zeros. Because of
singularities in the Green's functions at $z=0$, we use
$\sigma^B(x,y,z)$ with $z=851$ m, the shallowest level of our
calculations, in construction of the horizontal loading functions in
Equations \ref{eqn:f_hor_xz} and \ref{eqn:f_hor_yz}. Convolutions were
computed using a 2D fast Fourier transform. All calculations were
implemented in Python (v. 2.7.3) using IPython \citep{perez2007ipython},
NumPy (v. 1.7) \citep{oliphant2007numpy} and Pandas (v. 12)
\citep{mckinney2010}; additional statistical analysis was performed with
StatsModels \citep{seabold2010}. We created an open-source Python package
to calculate topographic stresses in a reasonably automated way, which
is available at \url{https://github.com/cossatot/halfspace}. The package
is being expanded to encompass a wide range of elastic stress and strain
solutions as time permits. All data and scripts for this particular
project are available at
\url{https://github.com/cossatot/wenchuan_topo_stress}.

@@

Topographic stresses on the Wenchuan faults are calculated on point sets
representing the faults taken from coseismic slip models. We use the
coseismic slip models of \citet{shen2009},
\citet{feng2010}, \citet{qi2011}, \citet{zhang2011},
and \citet{fielding2013}, and discard
points above 851 m below sea level, as this is above the depth at which
we compute $G^C(x,y,z)$. The six stress tensor components calculated at
the regular grid points are linearly interpolated to points describing
the faults. Because the fault points are completely surrounded by the
grid nodes at which topographic stresses were calculated and those nodes
are spaced \textless{}1 km apart, the fault points cannot be more than a
few hundred meters from the nearest grid node, so a higher order
interpolation is not necessary. We then project the topographic stress
tensor to fault normal stress, $\sigma_n^M$, down-dip shear stress,
$\tau_d^M$ and strike-slip shear stress, $\tau_s^M$ at each point in
describing the fault geometry.

@@

where $M$ is described above, $T$ is the tectonic stress tensor, and $L$
is the lithostatic stress tensor. $L$ is isotropic, with diagonal
components equal to $\rho g z$. We assume that $T$ is latterally
homogeneous and only has horizontal stress components (i.e.,
$T_{xz} = T_{yz} = T_{zz} = 0$, with $T_{xx}$, $T_{yy}$ and $T_{xy}$
non-zero). We further assume that $T$ increases linearly with depth so
that the entire upper crust is near the critical failure envelope
\citep[e.g.,][]{townend2000}, are thus we parameterize the components of $T$
as scalars multiplied by lithostatic pressure, denoted as $T^\prime$

@@

We define a joint posterior, $p_{J}(T^\prime | D)$, by the samples that 
are common to the individual posteriors estimated from each slip model.
Unsuprisingly, given the broad similarity between the posteriors from
the various slip models, $p_{J}(T^\prime_{\mathrm{max}} | D)$ is not
substantially different from any of the constituent model posteriors.
$p_{J}(T^\prime_{\mathrm{max}} | D)$ has a somewhat more well-defined
mode at $T^\prime_{\mathrm{max}} \approx 0.6$.

@@

The maximum tectonic stress, $T_{\mathrm{max}}$, is consistently
oriented roughly E-W in our results. This orientation is oblique to the
Longmen Shan, which produces oblique (right-lateral and reverse sense)
shear on the Beichuan fault. $T_{\mathrm{min}}$ is \textasciitilde{}N-S
oriented, and is only slightly larger than lithostatic pressure. This
stress configuration is compatible with the observed kinematics of the
Wenchuan faults, and in close agreement pre-earthquake stress
measurements near the rupture zone (Figure \ref{fig:lms_stress_map}),
mostly from borehole breakout data from 2-5 km depth
\citep{heidbach2009}, which is the zone of maximum slip in the coseismic
slip models. It is somewhat discrepant with stress orientations
estimated at \textasciitilde{}800 m depth adjacent to the Beichuan fault
in the WFSD-1 borehole of the Wenchuan Earthquake Fault Scientific
Drilling Project several years after the 2008 earthquake \citep{cui2014},
which show $\sigma_{H{\mathrm{\max}}}$ to be more orthogonal to the fault
trace, suggesting that much of the right-lateral component of shear
stress was released during the earthquake. These results are similar to
the orientations of total stress obtained by 
\citet{medinaluna2013}, although they were unable to constrain the
magnitudes of stresses.

>>>>

\begin{article} \section{Introduction}\label{introduction}
Stress is of fundamental importance to many processes in the earth. Both the
isotropic and deviatioric components of stress exert control on the deformation
state of the earth at any point in the brittle regime. However, unlike other
fundamental quantities such as temperature, stress is typically difficult to
measure \emph{in situ}, without drilling-based techniques. Therefore, stress is
often treated in a semi-quantitative manner, with an emphasis on directions and
relative magnitudes of the principal stresses, either locally or regionally
\citep[e.g.,][]{angelier1994}. These estimates of stress are commonly derived
from strain, for example from studies of earthquake focal mechanisms
\citep[e.g.,][]{michael1987} or of fault slip data \citep[e.g.,][]{reches1987,
medinaluna2013}.

In areas of substantial relief, high terrain and steep slopes generate large
stresses in the crust beneath and adjacent to the high topography
\citep{jeffreys1924, coblentz1996}. Because of the irregularity of topography
in mountainous regions, the stresses produced by topography are also
heterogeneous, and may play a prominent role in local or regional deformation,
particularly if the region is tectonically active.  For example, shear and
normal stresses on a fault from topographic loading may push a particular fault
closer to or farther from failure, or reorient the
net shear stress direction on a fault. These effects may affect the
localization or deformational style in a region. Furthermore, heterogeneous
topographic stresses on a particular fault may affect the way earthquake
ruptures propagate across the fault plane. Despite this, the degree to which
topographic stresses affect faulting has received little direct study.

@@

Aspects of topographic stresses and their relevance to tectonics have been
studied for some time. \citet{jeffreys1924} noted that the presence of high
mountains is evidence that the Earth's crust can support significant
heterogeneous differential stresses over long time periods.
\citet{dalmayrac1981} and \citet{molnar1988} discussed how extensional
deformation in the high parts of orogens that is temporally coincident with
contractional deformation at the low-elevation margins of the orogen may be
explained by a spatially invariant, depth-integrated horizontal compressive
stress with spatially varying vertical stresses caused by topography.
\citet{richardson1994} exploited this relationship in the central Andes to
estimate horizontal tectonic stresses through finite element modeling, and more
recently \citet{copley2009} performed similar work in Albania, and
\citet{fialko2005} investigated the role of vertical topographic stresses on
fault rotation in Southern California.
\citet{bollinger2004} showed how increased normal stress on the Main Himalayan
thrust due to loading of the high Himalayan massifs locally suppressed
microseismicity and increased fault locking. Meade and Conrad \citet{meade2008}
demonstrated that the increased weight of the uplifting Andes influenced the
Nazca-South America convergence rate.

@@

Topography was taken from the CGIAR-CSI v.4 release
\citep{jarvis2008srtm} of the Shuttle Radar Topographic Mission
\citep{farr2007srtm} Digital Elevation Model (DEM) at 1 km nominal
resolution. The DEM was projected from native WGS84 geographic
coordinates to UTM zone 48N, decreasing the nominal horizontal
resolution to 851 m. We assume a Poisson ratio of 0.25, following
receiver function studies suggesting values of ca. 0.24-0.26 throughout
central and western China \citep{chen2010receiverfn} (we have tested a Poisson
ratio of 0.28, which is on the higher end of values for intermediate rock
compositions \citep{zandt_ammons1995}, and found the results to vary by a few
percent). Green's
functions for the Boussinesq and Cerruti point-source solutions were
calculated at regular points in large 2-D grid at each depth
considered, with the point-source centered in the grid (see Table
\ref{table:convo_params} for model parameters). A mask was applied to
each of the discretized Green's functions such that values outside a 
radius (i.e. the `corners' of the array) were set to zero, yielding a
circular array. The size of the
grid was chosen to be quite large to incorporate potential contributions
from the elevated topography throughout eastern Tibet. So that the
Green's functions and the topography were discretized on the same size
grid, we pad the Green's function array with zeros. Because of
singularities in the Green's functions at $z=0$, we use
$\sigma^B(x,y,z)$ with $z=851$ m, the shallowest level of our
calculations, in construction of the horizontal loading functions in
Equations \ref{eqn:f_hor_xz} and \ref{eqn:f_hor_yz}. Convolutions were
computed using a 2D fast Fourier transform. All calculations were
implemented in Python (v. 2.7.3) using IPython \citep{perez2007ipython},
NumPy (v. 1.7) \citep{oliphant2007numpy} and Pandas (v. 12)
\citep{mckinney2010}; additional statistical analysis was performed with
StatsModels \citep{seabold2010}. We created an open-source Python package
to calculate topographic stresses in a reasonably automated way, which
is available at \url{https://github.com/cossatot/halfspace}. The package
is being expanded to encompass a wide range of elastic stress and strain
solutions as time permits. All data and scripts for this particular
project are available at
\url{https://github.com/cossatot/wenchuan_topo_stress}.

@@

Topographic stresses on the Wenchuan faults are calculated on point sets
representing the faults taken from coseismic slip models. We use six models,
those of \citet{shen2009}, \citet{feng2010}, \citet{zhang2011},
\citet{fielding2013} and two from \citet{qi2011} (these two models share a
fault geometry but use different smoothing functions for the slip
distribution). These vary somewhat in the fault geometry and specifics of the
slip distribution, but nonetheless are of high resolution, are properly
georeferenced. By using a suite of models in our calculations instead of a
single model, we evaluate results that are persistent in most or all of the
models to be more robust, and other results can be more confidently linked to
specifics of the model geometry or slip distribution.

In general, the models share a common geometry and pattern of slip distribution
above 10 km or so; i.e.\ the locations of high and low slip patches and the slip
rake are similar in all of the models. Though some of models use discrete fault
segments for the Beichuan fault, whereas others use a single continuous fault,
the fault planes are essentially collocated. However, there is significant
variability in the magnitude of slip in the different models; for example, the
maximum slip magnitude in the Qi `rough' model is about twice that of the
Feng model. Large differences also exist in the deeper geometries of the
models: the Qi, Fielding and Shen models all have horizontal or subhorizontal
thrust flats at or below 15 km, though only minor slip is assumed to have
occurred on these fault segments. In contrast, the Feng and Zhang models are
planar to their lower end at 25-30 km.

In our stress calculations, we discard
points above 851 m below sea level, as this is above the depth at which
we compute $G^C(x,y,z)$. The six stress tensor components calculated at
the regular grid points are linearly interpolated to points describing
the faults. Because the fault points are completely surrounded by the
grid nodes at which topographic stresses were calculated and those nodes
are spaced \textless{}1 km apart, the fault points cannot be more than a
few hundred meters from the nearest grid node, so a higher order
interpolation is not necessary. We then project the topographic stress
tensor to fault normal stress, $\sigma_n^M$, down-dip shear stress,
$\tau_d^M$ and strike-slip shear stress, $\tau_s^M$ at each point in
describing the fault geometry.

@@

where $M$ is described above, $T$ is the tectonic stress tensor, and $L$
is the lithostatic stress tensor. $L$ is isotropic, with diagonal
components equal to $\rho g z$. We assume that $T$ is laterally
homogeneous and only has horizontal stress components (i.e.,
$T_{xz} = T_{yz} = T_{zz} = 0$, with $T_{xx}$, $T_{yy}$ and $T_{xy}$
non-zero). The expanded stress tensor is:

\begin{equation}
S = \begin{bmatrix}
  M_{xx} + T_{xx} + L_{xx} & M_{xy} + T_{xy} &  M_{xz} \\
	M_{xy} + T_{xy} &  M_{yy} + T_{yy} + L_{yy} & M_{yz} \\
	M_{xz}     &  M_{yz}  &  M_{zz} + L_{zz}
	\end{bmatrix} \; .
\label{eqn:stress_tensor}
\end{equation}

We further assume that $T$ increases linearly with depth so
that the entire upper crust is near the critical failure envelope
\citep[e.g.,][]{townend2000}, and thus we parameterize the components of $T$
as scalars multiplied by $\rho g z$, denoted as $T^\prime$. Therefore, if
$T\prime_{xx} = 0.1$, at some point just below 1 km, $L=27$ MPa, so
$S_{xx} = M_{xx}$+ 27 MPa + 2.7 MPa.

@@

We define a joint posterior, $p_{J}(T^\prime | D)$, by the samples that 
are common to the individual posteriors estimated from each slip model.
Unsuprisingly, given the broad similarity between the posteriors from
the various slip models, $p_{J}(T^\prime_{\mathrm{max}} | D)$ is not
substantially different from any of the constituent model posteriors.
$p_{J}(T^\prime_{\mathrm{max}} | D)$ has a somewhat more well-defined
mode at $T^\prime_{\mathrm{max}} \approx 0.6$. It is also apparent in
Figure \ref{fig:joint_posteriors}~b that regardless of the magnitude of
$T^\_{\mathrm{max}$, $p_{J}(T^\prime_{\mathrm{min}} | D)$ is
consistently 0--0.6 $p_{J}(T^\prime_{\mathrm{max}} | D)$, with a mode
of about 0.3; this indicates that the relative magnitude of the
tectonic stresses has a substantial influence on the rake of the maximum
shear stress resolved on the fault.

@@

The maximum tectonic stress, $T_{\mathrm{max}}$, is consistently
oriented roughly E-W in our results. This orientation is oblique to the
Longmen Shan, which produces oblique (right-lateral and reverse sense)
shear on the Beichuan fault. $T_{\mathrm{min}}$ is \textasciitilde{}N-S
oriented, and is only slightly larger than lithostatic pressure. This
stress configuration is in close agreement with pre-earthquake stress
measurements near the rupture zone (Figure \ref{fig:lms_stress_map}),
mostly from borehole breakout data from 2-5 km depth
\citep{heidbach2009}, which is the zone of maximum slip in the coseismic
slip models. It is somewhat discrepant with stress orientations
estimated at \textasciitilde{}800 m depth adjacent to the Beichuan fault
in the WFSD-1 borehole of the Wenchuan Earthquake Fault Scientific
Drilling Project several years after the 2008 earthquake \citep{cui2014},
which show $\sigma_{H{\mathrm{\max}}}$ to be more orthogonal to the fault
trace, suggesting that much of the right-lateral component of shear
stress was released during the earthquake. These results are similar to
the orientations of total stress obtained by 
\citet{medinaluna2013}, although they were unable to constrain the
magnitudes of stresses.
